\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-11-23 Manuscript/manuscript.tex
smoothed world championship data

This paper has two primary objectives. First, we investigate whether 
reaction times at the 2022 World Championships were significantly 
different from other competitions, focusing on athletes who competed 
in multiple events. Using a matched-pairs design, we compare reaction 
times across the 2022, 2019, and 2023 World Championships, as well as 
the 2022 national-level competitions. This approach isolates the 
effect of the competition year while controlling for individual 
performance. The second objective is to assess the 
appropriateness of the 0.1-second threshold by modeling reaction 
times from World Championships held from 1999 onward. We use a gamma 
linear mixed-effects model with random effects for both venue and heat. 
\eds{edit previous as needed for new GAMLSS analysis}
This model estimates the probability of reaction times falling below 
the threshold, helping evaluate the consistency of reaction times and 
whether the current threshold is statistically justified.

The rest of this paper is organized as follows. Section~\ref{sec:Data} 
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the methods for
objective~1, analyzing differences in reaction times across
competitions, including the were collected. 
\eds{end of previous sentence does not make sense; some text may have been 
mistakenly deleted?}
Section~\ref{sec:gamlss}
details the methods for objective~2, the generalized additive model
for location, scale, and shape (GAMLSS) to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the 
paper’s impact, limitations, and potential for future research.
All data and code for our analysis 
are provided in the supplementary materials.

@@

The motivation for including data beyond the World Championships arose from 
an exploratory analysis of how United States athletes performed at the 2022 
United States Track and Field Championships, held from June 23--26, 2022, at 
Hayward Field in Eugene, Oregon. Since the same venue hosted the 2022 World 
Championships in August, this provided a unique opportunity to establish a 
baseline for reaction times.

@@

The final dataset consisted of reaction times from athletes who competed 
in both national and international competitions. Reaction times from 
preliminary heats, semifinals, and finals were included to ensure 
sufficient data for analysis. Excluding preliminary heats would have 
significantly reduced the number of athletes and clusters. Each
athlete was considered as a cluster, with observed reaction times
from both the `treatment' group (2022 World Championships) and
the `control' group (national competition) within the cluster. Cluster 
sizes ranged from three to six, with a median size of four. Because gender 
is known to influence reaction times \citep{lipps2011implications}, we
prepared data for men and women separately, resulting in 80 reaction
times from 17 athletes for each
gender. This setup forms the basis for the rank-based comparison for
clustered data with subunit-level grouping, which accounts for the
within-athlete dependence inherent in this data structure. The top
panel of Figure!~\ref{fig:RankScatterplots} shows the reaction times
of those who competed both at national competitions and the 2022 World
Championships.

@@

\eds{I also think we need to comment on the men and women's data. But within row
of the figure, don't the distributions appear similar?  If that is so, would a 
reviewer ask why we didn't pool the data for men and women?}

\subsection{World Championships 1999--2023}\label{sec:dataworld}

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2023.
We focus on the reaction times during the 
semifinal and final heats only, as reaction times from preliminary heats are 
often not as fast as those in later heats \citep[e.g.,][]{zhang2021correlation}. 
For analysis purposes, we will pool reaction times of men's final and 
men's semifinal heats together which increases the sample size. This is 
advantageous because in some years there are very few finals observations, such
as in 2022, when there were only five data points due to two disqualifications 
and one athlete who did not compete. This is the data that we will reference
throughout the rest of the paper unless otherwise noted. We also consider data 
that excludes 2022 to see how our analysis differs based on the exclusion of one
particular year of interest.

The data under consideration is summarized using side-by-side boxplots in 
Figure~\ref{fig:Boxplot}. It is clear that the reaction times in the 2022
boxplot are lower (indicating faster reaction times) than in many of the other
years. The median reaction time in 2022 was 0.129 seconds,  compared to 0.156
seconds, which is what \citet{brosnan2017effects} found when looking at data
from 1999 to 2014. We can also see from Figure~\ref{fig:Boxplot} that the 
reaction times clearly differ from year to year. The venue of the World 
Championships changes each year, so weather or climate related factors such as 
humidity, precipitation, elevation, may indeed play a role, but additionally,
both the technology and false start penalization rules have changed over the 
study period.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{Boxplot}
  \caption{The reaction times from 1999 to 2023 for the men's 110 meter hurdle
  and 100 meter dash.}
  \label{fig:Boxplot}
\end{figure}
\of{Do We want 2 boxplot figures: 1 hurdles, 1 dash or combine them}

From 2007 to 2009, the IAAF (the former name for World Athletics, and the 
governing body for the World Track and Field Championships) instituted a rule 
change that allowed one warning false start before a sprinter was disqualified 
\citep{iaaf2009falsestart}. There were 18 male and 7 female false starts at the 
2007 World Championships, and 18 male and 7 female false starts at the 2009 World 
Championships. Starting in 2011, the rule was scrapped, and the old policy which
automatically disqualifies runners who false start was reinstated. This was 
desirable for World Athletics as false starts can make already lengthy track 
meets more tedious, both for athletes and viewers watching the television 
broadcast. By returning to the harsher policy and cracking down on false starts,
World Athletics had reduced men's false starts by two thirds in 2011 (6 male and
4 female disqualifications) \citep{iaaf2009falsestart}. \citet{haugen2013effect}
examined the effect of different false start rules that IAAF has imposed from 
1997 to 2009 and found statistically significant improvement in reaction times 
under more lenient rules.

>>>>

This paper has two primary objectives. First, we investigate whether 
reaction times at the 2022 World Championships were significantly 
different from other competitions, focusing on athletes who competed 
in multiple events. Using a matched-pairs design, we compare reaction 
times across the 2022, 2019, and 2023 World Championships, as well as 
the 2022 national-level competitions. This approach isolates the 
effect of the competition year while controlling for individual 
performance. The clustered nature of the data, with the goal of
assessing differences across events within-athletes, were analyzed
with rank-based comparison approach for clustered data
\citep{datta2005rank}. The second objective is to assess the 
appropriateness of the 0.1-second threshold by modeling reaction 
times from World Championships held from 1999 onward. We use a
generalized gamma distribution with random effects for both venue and
heat in the framework of the generalized additive model for location,
scale, and shape (GAMLSS) \citep{stasinopoulos2024generalized}.
This model estimates the probability of reaction times falling below 
the threshold, helping evaluate the consistency of reaction times and 
whether the current threshold is statistically justified.

The rest of this paper is organized as follows. Section~\ref{sec:Data} 
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the rank-based comparison
methods for objective~1, rank-based comparison for clustered data with
subunit-level gruping to analyze differences in reaction times
across different competitions. Section~\ref{sec:gamlss} details the
methods for objective~2, the GAMLSS to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the 
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the supplementary
materials.

@@

For assessing the abnormality of the 2022 World Championship, 
the motivation for including data beyond the World Championships arose from 
an exploratory analysis of how United States athletes performed at the 2022 
United States Track and Field Championships, held from June 23--26, 2022, at 
Hayward Field in Eugene, Oregon. Since the same venue hosted the 2022 World 
Championships in August, this provided a unique opportunity to establish a 
baseline for reaction times.

@@

The final dataset consisted of reaction times from athletes who competed 
in both national and international competitions. Reaction times from 
preliminary heats, semifinals, and finals were included to ensure 
sufficient data for analysis. Excluding preliminary heats would have 
significantly reduced the number of athletes and clusters. Each
athlete was considered as a cluster, with observed reaction times
from both the `treatment' group (2022 World Championships) and
the `control' group (national competition) within the cluster. Cluster 
sizes ranged from three to six, with a median size of four. Because gender 
is known to influence reaction times \citep{babicc2009reaction,
  lipps2011implications}, we prepared data for men and women
separately, resulting in 80 reaction times from 17 athletes for each
gender. This setup forms the basis for the rank-based comparison for
clustered data with subunit-level grouping, which accounts for the
within-athlete dependence inherent in this data structure. The top
panel of Figure!~\ref{fig:RankScatterplots} shows the reaction times
of those who competed both at national competitions and the 2022 World
Championships.

@@

\eds{I also think we need to comment on the men and women's data. But within row
of the figure, don't the distributions appear similar?  If that is so, would a 
reviewer ask why we didn't pool the data for men and women?}
\jy{Let's put the pooled analysis in the supplement.}

\subsection{Data from World Championships 1999--2023}
\label{sec:dataworld}

The data for evaluating the appropriateness of the 0.1-second
threshold was obtained from World Athletics and covers the men's
110-meter hurdles and 100-meter dashes from 1999 to 2023. Due to
possible gender differences \citep{babicc2009reaction,
  lipps2011implications}, data for women's 100-meter hurdles and
100-meter dashes and their analyses were relegated to the
supplementary material. We focus on the reaction times recorded during
semifinal and final heats only, as reaction times from preliminary
heats are often not as fast as those in later heats
\citep[e.g.,][]{collet1999strategic, tonnessen2013reaction,
  brosnan2017effects, zhang2021correlation}. For analysis purposes, we
pooled reaction times from semifinal and final heats to increase sample 
size, which is particularly important for years with limited final heat 
observations. For example, in 2022, only five data points were available 
from the final heat due to two disqualifications and one athlete not 
competing. Unless otherwise noted, this pooled dataset forms the basis 
for our analysis throughout the paper. Additionally, we consider datasets 
that exclude 2022 to assess how our findings might differ when excluding 
this year of interest.

\jy{have the numbers in this paragraph been updated with the
  additional 100-dash data?}
The data is summarized in Figure~\ref{fig:Boxplot}, which presents a
sequence of boxplots of reaction times from 1999 to 2023. It is evident 
that reaction times in 2022 were notably faster, with a median reaction 
time of 0.129 seconds compared to the 0.156 seconds observed in earlier 
studies, such as \citet{brosnan2017effects} for data spanning 1999 to 2014. 
Figure~\ref{fig:Boxplot} also highlights year-to-year variability in 
reaction times, likely influenced by changes in the championship venue 
and environmental conditions such as humidity, precipitation, and 
elevation. Furthermore, advancements in technology and alterations to 
false start rules during the study period may have played a role in these 
variations \citep{willwacher2013novel}.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{Boxplot}
  \caption{The reaction times from 1999 to 2023 for the men's 110 meter hurdle
  and 100 meter dash.}
  \label{fig:Boxplot}
\end{figure}

Between 2007 and 2009, World Athletics (formerly the IAAF) allowed one 
false start warning before disqualifying a sprinter \citep{iaaf2009falsestart}. 
This lenient rule led to 18 male and 7 female false starts at both the 2007 
and 2009 World Championships. In 2011, this rule was replaced with the stricter 
policy of automatic disqualification for false starts, aimed at reducing the 
delays caused by repeated warnings. This change reduced men’s false starts 
by two-thirds in 2011, with only six male and four female disqualifications 
\citep{iaaf2009falsestart}. \citet{haugen2013effect} demonstrated that more 
lenient false start rules significantly improved reaction times during the 
1997–2009 period, suggesting that rule changes over the study period may 
have contributed to variations in reaction times across years.
